---
title: "geeCRT package"
author: "Hengshi Yu, Fan Li, Paul Rathouz, Elizabeth L. Turner, John Preisser"
date: "`r Sys.Date()`"
bibliography: references.bib
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{geeCRT package for the design and analysis of cluster randomized trials}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
```{r start,echo=FALSE,results="hide"}

library(geeCRT)

```


## Overview
geeCRT is an R package for implementing the bias-corrected generalized estimating equations in analyzing cluster randomized trials.

Population-averaged models have been increasingly used in the design and analysis of cluster randomized trials (CRTs). To facilitate the applications of population-averaged models in CRTs, we implement the generalized estimating equations (GEE) and matrix-adjusted estimating equations (MAEE) approaches to jointly estimate the marginal mean models correlation models both for general CRTs and stepped wedge CRTs.

Despite the general GEE/MAEE approach, we also implement a fast cluster-period GEE method specifically for stepped wedge CRTs with large and variable cluster-period sizes. The individual-level GEE/MAEE approach becomes computationally infeasible in this setting due to inversion of high-dimensional covariance matrices and the enumeration of a high-dimensional design matrix for the correlation estimation equations. The package gives a simple and efficient estimating equations approach based on the cluster-period means to estimate the intervention effects as well as correlation parameters.

In addition, the package also provides functions for generating correlated binary data with specific mean vector and correlation matrix based on the multivariate probit method [@emrich1991method] or the conditional linear family method ([@qaqish2003family]). These two functions facilitate generating correlated binary data in future simulation studies. 

## `geemaee()` example: matrix-adjusted GEE for estimating the mean and correlation parameters in CRTs

The `geemaee()` function implements the matrix-adjusted GEE or regular GEE developed for analyzing cluster randomized trials (CRTs). It provides valid estimation and inference for the treatment effect and intraclass correlation parameters within the population-averaged modeling framework. The program allows for flexible marginal mean model specifications. The program also offers bias-corrected intraclass correlation coefficient (ICC) estimates as well as bias-corrected sandwich variances for both the treatment effect parameter and the ICC parameters. The technical details of the matrix-adjusted GEE approach are provided in Preisser et al. (2008) and Li et al. (2018).

For the individual-level data, we use the `geemaee()` function to estimate the marginal mean and correlation parameters in CRTs. We use two simulated stepped wedge CRT datasets with true nested exchangeable correlation structure to illustrate the `geemaee()` function examples. We first create an auxiliary function `createzCrossSec()` to help create the design matrix for the estimating equations of the correlation parameters. We then collect design matrix `X` for the mean parameters with five period indicators and the treatment indicator.

```{r createz, fig.keep="all", fig.width = 7, fig.height=4}

createzCrossSec = function (m) {

    Z = NULL
    n = dim(m)[1]
    
    for (i in 1:n) {
        
        alpha_0 = 1; alpha_1 = 2; n_i = c(m[i, ]); n_length = length(n_i)
        POS = matrix(alpha_1, sum(n_i), sum(n_i))
        loc1 = 0; loc2 = 0
        
        for (s in 1:n_length) {
            
            n_t = n_i[s]; loc1 = loc2 + 1; loc2 = loc1 + n_t - 1
            
            for (k in loc1:loc2) {

                for (j in loc1:loc2) {

                    if (k != j) { POS[k, j] = alpha_0 } else { POS[k, j] = 0 }}}}

        zrow = diag(2); z_c = NULL
        
        for (j in 1:(sum(n_i) - 1)) { 

            for (k in (j + 1):sum(n_i)) {z_c = rbind(z_c, zrow[POS[j,k],])}}
        
        Z = rbind(Z, z_c) }

    return(Z)}
``` 

We implement the `geemaee()` function on both the continuous outcome and binary outcome, and consider both matrix-adjusted estimating equations (MAEE) with `alpadj = TRUE` and uncorrected generalized estimating equations (GEE) with `alpadj = FALSE`. For the `shrink` argument, we use the `"ALPHA"` method to tune step sizes and focus on using estimated variances in the correlation estimating equations rather than using unit variances by specifying `makevone = FALSE`. 


```{r cvrall, fig.keep="all", fig.width = 7, fig.height=4}

 Design_result <- cvrall(clustername = Dickinson_design$county,
                  balancemetric = "l2",
                  x = data.frame(Dickinson_design[ , c("location", "inciis",
                      "uptodateonimmunizations", "hispanic", "incomecat")]),
                  ntotal_cluster = 16,
                  ntrt_cluster = 8,
                  categorical = c("location", "incomecat"),
                  ###### Option to save the constrained space #####
                  # savedata = "dickinson_constrained.csv",
                  bhist = TRUE,
                  cutoff = 0.1,
                  seed = 12345)
 

```


The we had the following output: 
```{r set-options1, echo=FALSE, fig.keep="all", fig.width = 7, fig.height=4}
options(width = 100)
 
```

```{r , fig.keep="all", fig.width = 7, fig.height=4}
 # the balance metric used
 Design_result$balancemetric

 # the allocation scheme from constrained randomization
 Design_result$allocation
 
 # the histogram of the balance score with respect to the balance metric
 Design_result$bscores
 
 # the statement about how many clusters to be randomized to the intervention and the control arms respectively
 Design_result$assignment_message
 
 # the statement about how to get the whole randomization space to use in constrained randomization
 Design_result$scheme_message
 
 # the statement about the cutoff in the constrained space
 Design_result$cutoff_message
 
 # the statement about the selected scheme from constrained randomization
 Design_result$choice_message
 
 
 # the data frame containing the allocation scheme, the clustername as well as the original data frame of covariates
 Design_result$data_CR
 
 # the descriptive statistics for all the variables by the two arms from the selected scheme
 Design_result$baseline_table

 # the cluster pair descriptive, which is useful for valid randomization check
 Design_result$cluster_coin_des

 # the overall allocation summary
 Design_result$overall_allocations

```

From the output of `Design_result$baseline_table`, the selected scheme is able to properly balance the baseline values of the covariates. The selected scheme is shown in `Design_result$allocation`. 

## `cvrall()` example for stratified constrained randomization

User-defined weights can be used to induce stratification on one or more categorical variables. In the study presented by [@dickinson2015pragmatic], there are 8 `"Urban"` and 8 `"Rural"` counties. A user-defined weight of `1,000` is added to the covariate of `location`, while these weights for other covariates are all `1`. Intuitively, a large weight assigned to a covariate sharply penalizes any imbalance of that covariates, therefore including schemes that are optimally balanced with respect to that covariate in the constrained randomization space. In practice, the resulting constrained space approximates the stratified randomization space on that covariate. In our illustrative data example, since half of the counties are located in rural areas, perfect balance is achieved by considering constrained randomization with the large weight for `location` variable. Alternatively, the option of `stratify` is able to perform the equivalent stratification on the stratifying variables specified.

```{r cvrallst1, fig.keep="all", fig.width = 7, fig.height=4}
# Stratification on location, with constrained randomization on other specified covariates

Design_stratified_result1 <- cvrall(clustername = Dickinson_design$county,
                                     balancemetric = "l2",
                                     x = data.frame(Dickinson_design[ , c("location", "inciis", 
                                                                          "uptodateonimmunizations", 
                                                                          "hispanic", "incomecat")]),
                                     ntotal_cluster = 16,
                                     ntrt_cluster = 8,
                                     categorical = c("location", "incomecat"),
                                     weights = c(1000, 1, 1, 1, 1),
                                     cutoff = 0.1,
                                     seed = 12345) 

```

```{r , fig.keep="all", fig.width = 7, fig.height=4}
Design_stratified_result1$baseline_table

```

```{r cvrallst2, fig.keep="all", fig.width = 7, fig.height=4}
# An alternative and equivalent way to stratify on location

Design_stratified_result2 <- cvrall(clustername = Dickinson_design$county,
                                     balancemetric = "l2",
                                     x = data.frame(Dickinson_design[ , c("location", "inciis",
                                                                          "uptodateonimmunizations", 
                                                                          "hispanic", "incomecat")]),
                                     ntotal_cluster = 16,
                                     ntrt_cluster = 8,
                                     categorical = c("location", "incomecat"),
                                     stratify = "location",
                                     cutoff = 0.1,
                                     seed = 12345, 
                                     check_validity = TRUE)

```




```{r , fig.keep="all", fig.width = 7, fig.height=4}
Design_stratified_result2$baseline_table

```
The results from `Design_stratified_result1$baseline_table` and `Design_stratified_result2$baseline_table` are the same. The final selected scheme from cvrall() now has 4 `"Urban"` counties in both arms. The `location` covariate has been stratified for the randomization for the randomization through the `weights` or `stratify` argument in the ` cvrall()` function. 


## `cvrcov()` example for covariate-by-covariate constrained randomization


For the covariate-by-covariate randomization, we used the `cvrcov()` function to randomize 8 out of the 16 counties into the practice-based. For the definition of the whole randomization space, if the total number of all possible schemes is smaller than `100,000`, we enumerate all the schemes as the whole randomization space. Otherwise, we simulate `100,000` unique schemes. Location has `"Rural"` and `"Urban"`. The level of `"Rural"` was then kept as 1 in `cvrcov()` and `"Urban"` is 0. Then we constrained the randomization space to have the schemes with absolute total difference of location be smaller than or equal to `5`, absolute mean difference of percentages of children ages 19-35 months in the CIIS less than or equal to `0.5` fraction of the overall mean, and absolute mean difference of income to be less than or equal to the `0.4` fraction of the overall mean. Finally, a randomization scheme is sampled from the constrained space. 

We saved the constrained randomization space in a CSV file in `"dickinson_cov_constrained.csv"`, the first column of which is an indicator variable of the finally selected scheme (`1`) or not (`0`). 



```{r cvrcov, fig.keep="all", fig.width = 7, fig.height=4}

# change the categorical variable of interest to have numeric representation
Dickinson_design_numeric <- Dickinson_design
Dickinson_design_numeric$location = (Dickinson_design$location == "Rural") * 1

Design_cov_result <- cvrcov(clustername = Dickinson_design_numeric$county,
                            x = data.frame(Dickinson_design_numeric[ , c("location", "inciis", 
                                                                          "uptodateonimmunizations", 
                                                                          "hispanic", "income")]),
                            ntotal_cluster = 16,
                            ntrt_cluster = 8,
                            constraints = c("s5", "mf.5", "any", "any", "mf0.4"), 
                            categorical = c("location"),
                            ###### Option to save the constrained space #####
                            # savedata = "dickinson_cov_constrained.csv",
                            seed = 12345, 
                            check_validity = TRUE)
 

```


The we had the following output: 
```{r set-options2, echo=FALSE, fig.keep="all", fig.width = 7, fig.height=4}
options(width = 100)
 
```

```{r , fig.keep="all", fig.width = 7, fig.height=4}


 # the allocation scheme from constrained randomization
 Design_cov_result$allocation
 

 # the statement about how many clusters to be randomized to the intervention and the control arms respectively
 Design_cov_result$assignment_message
 
 # the statement about how to get the whole randomization space to use in constrained randomization
 Design_cov_result$scheme_message
 

 # the data frame containing the allocation scheme, the clustername as well as the original data frame of covariates
 Design_cov_result$data_CR
 
 # the descriptive statistics for all the variables by the two arms from the selected scheme
 Design_cov_result$baseline_table

# the cluster pair descriptive, which is useful for valid randomization check
Design_cov_result$cluster_coin_des

# the overall allocation summary
Design_cov_result$overall_allocations

```

From the output of `Design_cov_result$baseline_table`, the selected scheme is able to properly balance the baseline values of the covariates. The selected scheme is shown in `Design_cov_result$allocation`. 


## Analysis: clustered permutation test

At the end of cluster randomized trials, individual outcomes are collected. Permutation test based on [@gail1996design] and [@li2016evaluation] is then applied to the continuous or binary outcome with some individual-level covariates. 

The permutation test is implemented in a two-step procedure. In the first step, an outcome regression model is fitted for response $Y_{ij}$ with covariates $\textbf{z}_{ij}$. This is done by fitting a linear regression model for continuous responses and a logistic regression model for binary responses [@gail1996design], ignoring the clustering of responses. The individual residual $r_{ij}=Y_{ij}-\hat{Y}_{ij}$ can be calculated from the predicted response for each individual by $\hat{Y}_{ij}$. In the second step, cluster-specific residual means are obtained as $\bar{r}_{i\cdot}=\sum_{j=1}^{m_i}r_{ij}/m_i$. The observed test statistic is then computed as $U=\frac{1}{n_T}\sum_{i=1}^nW_i\bar{r}_{i\cdot}-
\frac{1}{n_C}\sum_{i=1}^n(1-W_i)\bar{r}_{i\cdot}$, where $W_i=1$ if the $i$th cluster is assigned to the treatment arm and $W_i=0$ otherwise, and $n_T=\sum_{i=1}^nW_i$, $n_C=\sum_{i=1}^n(1-W_i)$ are the number of treated and control clusters.

Suppose there are $S$ randomization schemes in the constrained randomization space. To obtain the permutation distribution of the test statistic, we permute the labels of the treatment indicator according to the constrained randomization space, and compute a value of $U_s$ ($s=1,\ldots,S$). The collection of these values $\{U_s:s=1,\ldots,S\}$ forms the null distribution of the permutation test statistic. The p-value is then computed by $\text{p-value}=\frac{1}{S}\sum_{s=1}^S \mathbb{I}(|U_s|\geq |U|)$. 

The `cptest()` function in the `cvcrand` package is used to perform the permutation test for the intervention effect of cluster randomized trials. 

Each categorical variable is transformed into dummy variables to fit in the linear model or logistic regression for the permutation test. When transforming a categorical variable to dummy variables, the reference level will be dropped if the categorical variable is specified as a factor. Otherwise, the first level in the alphanumerical order will be dropped. Users can also specify a certain reference level for each categorical variable by manually coding dummy variables before running the `cptest()` function.

## `cptest()` example

Suppose that the researchers were able to assess 300 children in each cluster in a study presented by [@dickinson2015pragmatic], and the cluster randomized trial is processed with the selected randomization scheme from the example above of the `cvrall()` function. We expanded the values of the cluster-level covariates on the covariates' values of the individuals, according to which cluster they belong to. The correlated individual outcome of up-to-date on immunizations (`1`) or not (`0`) is then simulated using a generalized linear mixed model (GLMM) with a logistic link to induce correlation by including a random effect at the county level. The intracluster correlation (ICC) was set to be 0.01, using the latent response definition provided in [@eldridge2009intra]. This is a reasonable value for population health studies [@hannan1994parameters]. We simulated one data set, with the outcome data dependent on the county-level covariates used in the constrained randomization design and a positive treatment effect so that the practice-based intervention increases up-to-date immunization rates more than the community-based intervention. For each individual child, the outcome is equal to `1` if he or she is up-to-date on immunizations and `0` otherwise.

```{r, echo=FALSE, results='asis'}
knitr::kable(head(Dickinson_outcome, 10))
```

We used the `cptest()` function to process the clustered permutation test on the binary outcome of the status of up-to-date on immunizations. We input the file about the constrained space with the first column indicating the final scheme. The permutation test is on the continuous covariates of `"inciis"`, `"uptodateonimmunizations"`, `"hispanic"`, as well as categorical variables of `"location"` and `"incomecat"`. Location has `"Rural"` and `"Urban"`. The level of `"Rural"` was then dropped in `cptest()`. As income category has three levels of `"low"`, `"med"`, and `"high"`,  the level of `"high"` was dropped to create dummy variables according to the alphanumerical order as well.

```{r cptest, fig.keep="all", fig.width = 7, fig.height=4}
 Analysis_result <- cptest(outcome = Dickinson_outcome$outcome,
                           clustername = Dickinson_outcome$county,
                           z = data.frame(Dickinson_outcome[ , c("location", "inciis",
                               "uptodateonimmunizations", "hispanic", "incomecat")]), 
                            cspacedatname = system.file("dickinson_constrained.csv", package = "cvcrand"),                                 
                           outcometype = "binary",                                                      
                           categorical = c("location","incomecat"))


``` 

The result of `"cptest()"` includes the final scheme for the cluster randomized trial, the p-value from the permutation test as well as a statement about that p-value.

```{r cptestre, fig.keep="all", fig.width = 7, fig.height=4}
 Analysis_result 
``` 

From the p-value of `0.042` in `Analysis_result`, the probability of up-to-date on immunizations for the practice-based approach (`1`) is significantly different from that for the population-based approach (`0`). 

# Session Information

```{r info, results='markup', echo=FALSE}
sessionInfo()
```

# References
